\pkg{SWMPr}: An R Package for Retrieving, Organizing, and Analyzing Environmental Data for Estuaries

### 2015-11-17 Rjournal/beck.tex
journal resubmission files prepped

\abstract{
The System-Wide Monitoring Program (SWMP) was implemented in 1995 by the US National Estuarine Research Reserve System. This program has provided two decades of continuous monitoring data at over 140 fixed stations in 28 estuaries. Simple tools for processing and evaluating the increasing quantity of data provided by the monitoring network have prevented broad-scale comparisons between systems and, in some cases, simple trend analysis of water quality parameters at individual sites.  This article describes the \CRANpkg{SWMPr} package that provides several functions to facilitate data retrieval, organization, and analysis of time series data in the reserve estuaries.  Previously unavailable functions for estuaries are also provided to estimate rates of ecosystem metabolism using the open-water method.  The \CRANpkg{SWMPr} package has facilitated a cross-reserve comparison of water quality trends and links quantitative information with analysis tools that has use for more generic applications to environmental time series.
}

@@

\begin{example}
# import all paramaters for the station
# three most recent records
exdat <- all_params('apadbwq', Max = 3, trace = F)
exdat
##         datetimestamp temp f_temp spcond f_spcond sal f_sal do_pct
## 1 2015-08-18 07:00:00   28      0     34        0  21     0     77
## 2 2015-08-18 07:15:00   28      0     34        0  21     0     75
## 3 2015-08-18 07:30:00   28      0     33        0  21     0     82
##   f_do_pct do_mgl f_do_mgl depth f_depth ph f_ph turb f_turb chlfluor
## 1        0      5        0     2       0  8    0   13      0       NA
## 2        0      5        0     2       0  8    0    9      0       NA
## 3        0      6        0     2       0  8    0   11      0       NA
##   f_chlfluor level f_level cdepth clevel f_cdepth f_clevel
## 1         -2    NA      -1     NA     NA       -1         
## 2         -2    NA      -1     NA     NA       -1         
## 3         -2    NA      -1     NA     NA       -1
\end{example}

@@

Viewing the number of observations for each parameter that are assigned to a QAQC flag may be useful for deciding how to process the data with \code{qaqc}.  The \code{qaqcchk} function can be used to view this information.

@@

% table for analysis functions
%latex.default(to_tab[, "Description", drop = F], file = "", caption = "Analysis functions available from the \\pkg{SWMPr} package.  Full documentation for each function is in the help file (e.g., execute \\code{?aggreswmp} for individual functions or \\code{help.search(`analyze', package = `SWMPr')} for all).",     rowlabel = "Function", colheads = "Description", rowname = to_tab$Functions,     caption.loc = "top", col.just = c("p{3.5in}"), label = "tab:analyze",     table.env = FALSE, where = "!tbp")%
\begin{table}[!tbp]
\caption{Analysis functions available from the \pkg{SWMPr} package.  Full documentation for each function is in the help file (e.g., execute \code{?aggreswmp} for individual functions or \code{help.search(`analyze', package = `SWMPr')} for all).\label{tab:analyze}} 
\begin{center}
\begin{tabular}{lp{3.5in}}
\hline\hline
\multicolumn{1}{l}{Function}&\multicolumn{1}{c}{Description}\tabularnewline
\hline
\code{aggreswmp}&Aggregate \code{"swmpr"} objects for different time periods - years, quarters, months,  weeks, days, or hours.  The aggregation function defaults to the mean.\tabularnewline
\code{aggremetab}&Aggregate metabolism data from a \code{"swmpr"} object.  This is primarly used within \code{plot\_metab} but may be useful for simple summaries of daily metabolism data.\tabularnewline
\code{ecometab}&Estimate ecosystem metabolism for a combined water quality and weather dataset using the open-water method \citep{Odum56}.\tabularnewline
\code{decomp}&Decompose a \code{"swmpr"} time series into trend, seasonal, and residual components.  This is a simple wrapper to \code{decompose} \citep{Kendall83}.  Decomposition of monthly or daily trends is possible.\tabularnewline
\code{decomp\_cj}&Decompose a \code{"swmpr"} time series into grandmean, annual, seasonal, and events components.  This is a simple wrapper to \code{decompTs} in the \CRANpkg{wq} package \citep{Jassby14}.  Only monthly decomposition is possible.\tabularnewline
\code{hist}&Plot a histogram for a single variable.\tabularnewline
\code{lines}&Add lines to an existing plot created with \code{plot}.\tabularnewline
\code{map\_reserve}&Create a map of all stations in a reserve using the \CRANpkg{ggmap} package \citep{Kahle13}.\tabularnewline
\code{na.approx}&Linearly interpolate missing data (\code{NA} values) in a \code{"swmpr"} object.\tabularnewline
\code{overplot}&Plot multiple time series on the same y-axis.\tabularnewline
\code{plot}&Plot a univariate  time series for a \code{"swmpr"} object.\tabularnewline
\code{plot\_metab}&Plot ecosystem metabolism estimates after running \code{ecometab} on a combined \code{"swmpr"} object.\tabularnewline
\code{plot\_summary}&Create summary plots of seasonal/annual trends and anomalies for a single parameter.\tabularnewline
\code{smoother}&Smooth \code{"swmpr"} objects with a moving window average, passed to \code{filter}.\tabularnewline
\hline
\end{tabular}\end{center}

@@

# decomposition of chl
decomp_cj(dat, param = 'chla_n')
\end{example}
\begin{figure}[!h]

@@

## plot
plot_summary(dat, param = 'chla_n', years = c(2007, 2013))
\end{example}
\begin{figure}[!h]

@@

## plot
overplot(dat, select = c('depth', 'do_mgl', 'ph', 'turb'),
  subset = c('2013-01-01 0:0', '2013-02-01 0:0'), lwd = 2)
\end{example}
\begin{figure}[!h]

@@

Finally, the \code{map\_reserve} function can be used to create a map of stations at a reserve using the \CRANpkg{ggmap} package \citep{Kahle13}. The function uses Google maps of four types that can be set with the \code{map\_type} argument: terrain (default), satellite, roadmap, or hybrid.  The \code{zoom} argument can be chosen through trial and error depending on the spatial extent of the reserve.

\begin{example}
# plot the stations at Jacques Cousteau reserve
map_reserve('jac')
\end{example}
\begin{figure}[!h]

@@

Two \CRANpkg{shiny} web applications illustrate the improved ability to synthesize and evaluate multi-year time series using \pkg{SWMPr}.  The first application evaluates trends in SWMP data within and between sites using an interactive \pkg{leaflet} map (\citet{Cheng15}, Figure~\ref{fig:swmp_comp}): \href{https://beckmw.shinyapps.io/swmp_comp}{https://beckmw.shinyapps.io/swmp\_comp}.  Trends between reserves can be viewed using the map, whereas trends at individual sites can be viewed by clicking on a map location.  Site-level trends are shown below the map with a simple linear regression to show an increase or decrease in values over time, whereas trends between sites are shown on the map for each station as circles that identify the direction and significance of the trend.  The second application provides graphical summaries of water quality, weather, or nutrient station data at individual stations using the \code{plot\_summary} function: \href{https://beckmw.shinyapps.io/swmp_summary/}{https://beckmw.shinyapps.io/swmp\_summary}.  The output is identical to Figure~\ref{fig:summary_ex} with the addition of drop down menus to select the station, date range, and parameter for plotting.

\begin{figure}
\begin{center}
\includegraphics[width = \textwidth]{swmp_comp.pdf}
\caption{Online application for comparing trends in SWMP data parameters using an interactive map.  Link: \href{https://beckmw.shinyapps.io/swmp_comp}{https://beckmw.shinyapps.io/swmp\_comp}}
\label{fig:swmp_comp}
\end{center}
\end{figure}

>>>>

\abstract{
The System-Wide Monitoring Program (SWMP) was implemented in 1995 by the US National Estuarine Research Reserve System. This program has provided two decades of continuous monitoring data at over 140 fixed stations in 28 estuaries. However, the increasing quantity of data provided by the monitoring network has complicated broad-scale comparisons between systems and, in some cases, prevented simple trend analysis of water quality parameters at individual sites.  This article describes the \CRANpkg{SWMPr} package that provides several functions to facilitate data retrieval, organization, and analysis of time series data in the reserve estuaries.  Previously unavailable functions for estuaries are also provided to estimate rates of ecosystem metabolism using the open-water method.  The \CRANpkg{SWMPr} package has facilitated a cross-reserve comparison of water quality trends and links quantitative information with analysis tools that has use for more generic applications to environmental time series.
}

@@

\begin{example}
# import all paramaters for the station
# three most recent records
exdat <- all_params('apadbwq', Max = 3, trace = F)
exdat
##         datetimestamp temp f_temp spcond f_spcond sal f_sal do_pct
## 1 2015-11-03 11:15:00   26      0     45        0  29     0     78
## 2 2015-11-03 11:30:00   26      0     46        0  30     0     76
## 3 2015-11-03 11:45:00   26      0     46        0  30     0     75
##   f_do_pct do_mgl f_do_mgl depth f_depth ph f_ph turb f_turb chlfluor
## 1        0      5        0     2       0  8    0    2      0       NA
## 2        0      5        0     2       0  8    0    5      0       NA
## 3        0      5        0     2       0  8    0    5      0       NA
##   f_chlfluor level f_level cdepth clevel f_cdepth f_clevel
## 1         -2    NA      -1      2     NA        3         
## 2         -2    NA      -1      2     NA        3         
## 3         -2    NA      -1      2     NA        3
\end{example}

@@

SWMP data often contain observations above or below the detection limit for the sensor or laboratory method used to quantify the parameters.  For example, nutrient data exceeding the high sensor range are assigned a QAQC flag of -5, whereas data below the low sensor range are assigned a QAQC flag of -4.  The presence of censored data is non-trivial and can influence the types of analyses that are appropriate for a time series \citep{Helsel12}.  A detailed discussion of methods for evaluating censored data is beyond the scope of the manuscript and existing methods for R are provided by other packages \citep[e.g., \CRANpkg{cents},][]{McLeod14}.  However, the functions in \pkg{SWMPr} can be used to identify censored data based on the appropriate QAQC flag for a parameter.  Viewing this information can be helpful for determining how to further process the data with the \code{qaqc} function or alternative methods outside of \pkg{SWMPr}.  The \code{qaqcchk} function returns a \code{data.frame} of the number of observations for a parameter that are assigned to all QAQC flags, including those for censored data.  SWMP data should not be analyzed without viewing this information to determine an appropriate method to address data with questionable QAQC flags.

@@

% table for analysis functions
%latex.default(to_tab[, "Description", drop = F], file = "", caption = "Analysis functions available from the \\pkg{SWMPr} package.  Full documentation for each function is in the help file (e.g., execute \\code{?aggreswmp} for individual functions or \\code{help.search(`analyze', package = `SWMPr')} for all).",     rowlabel = "Function", colheads = "Description", rowname = to_tab$Functions,     caption.loc = "top", col.just = c("p{3.5in}"), label = "tab:analyze",     table.env = FALSE, where = "!tbp")%
\begin{table}[!tbp]
\caption{Analysis functions available from the \pkg{SWMPr} package.  Full documentation for each function is in the help file (e.g., execute \code{?aggreswmp} for individual functions or \code{help.search(`analyze', package = `SWMPr')} for all).\label{tab:analyze}} 
\begin{center}
\begin{tabular}{lp{3.5in}}
\hline\hline
\multicolumn{1}{l}{Function}&\multicolumn{1}{c}{Description}\tabularnewline
\hline
\code{aggreswmp}&Aggregate \code{"swmpr"} objects for different time periods - years, quarters, months,  weeks, days, or hours.  The aggregation function defaults to the mean.\tabularnewline
\code{aggremetab}&Aggregate metabolism data from a \code{"swmpr"} object.  This is primarly used within \code{plot\_metab} but may be useful for simple summaries of daily metabolism data.\tabularnewline
\code{ecometab}&Estimate ecosystem metabolism for a combined water quality and weather dataset using the open-water method \citep{Odum56}.\tabularnewline
\code{decomp}&Decompose a \code{"swmpr"} time series into trend, seasonal, and residual components.  This is a simple wrapper to \code{decompose} \citep{Kendall83}.  Decomposition of monthly or daily trends is possible.\tabularnewline
\code{decomp\_cj}&Decompose a \code{"swmpr"} time series into grandmean, annual, seasonal, and events components.  This is a simple wrapper to \code{decompTs} in the \CRANpkg{wq} package \citep{Jassby14}.  Only monthly decomposition is possible.\tabularnewline
\code{hist}&Plot a histogram for a single variable.\tabularnewline
\code{lines}&Add lines to an existing plot created with \code{plot}.\tabularnewline
\code{map\_reserve}&Create a map of all stations in a reserve using the \CRANpkg{ggmap} package \citep{Kahle13}.\tabularnewline
\code{na.approx}&Linearly interpolate missing data (\code{NA} values) in a \code{"swmpr"} object.\tabularnewline
\code{overplot}&Plot multiple time series in a \code{"swmpr"} object on the same y-axis.\tabularnewline
\code{plot}&Plot a univariate  time series for a \code{"swmpr"} object.\tabularnewline
\code{plot\_metab}&Plot ecosystem metabolism estimates after running \code{ecometab} on a combined \code{"swmpr"} object.\tabularnewline
\code{plot\_summary}&Create summary plots of seasonal/annual trends and anomalies for a single parameter.\tabularnewline
\code{smoother}&Smooth \code{"swmpr"} objects with a moving window average, passed to \code{filter}.\tabularnewline
\hline
\end{tabular}\end{center}

@@

# decomposition of chl
decomp_cj(dat, param = 'chla_n')
\end{example}

\begin{figure}[!h]

@@

## plot
plot_summary(dat, param = 'chla_n', years = c(2007, 2013))
\end{example}

\begin{figure}[!h]

@@

## plot
overplot(dat, select = c('depth', 'do_mgl', 'ph', 'turb'),
  subset = c('2013-01-01 0:0', '2013-02-01 0:0'), lwd = 2)
\end{example}

\begin{figure}[!h]

@@

Finally, the \code{map\_reserve} function can be used to create a map of stations at a reserve using the \CRANpkg{ggmap} package (Figure~\ref{fig:map_ex}, \citet{Kahle13}). The function uses Google maps of four types that can be set with the \code{map\_type} argument: terrain (default), satellite, roadmap, or hybrid.  The \code{zoom} argument can be chosen through trial and error depending on the spatial extent of the reserve.

\begin{example}
# plot the stations at Jacques Cousteau reserve
map_reserve('jac')
\end{example}

\begin{figure}[!h]

@@

Two \CRANpkg{shiny} web applications illustrate the improved ability to synthesize and evaluate multi-year time series using \pkg{SWMPr}.  The first application evaluates trends in SWMP data within and between sites using an interactive \pkg{leaflet} map (\citet{Cheng15}, Figure~\ref{fig:swmp_comp}): \href{https://beckmw.shinyapps.io/swmp_comp}{https://beckmw.shinyapps.io/swmp\_comp}.  Trends between reserves can be viewed using the map, whereas trends at individual sites can be viewed by clicking on a map location.  Site-level trends are shown below the map with a simple linear regression to show an increase or decrease in values over time, whereas trends between sites are shown on the map for each station as circles that identify the direction and significance of the trend.  More robust methods for evaluating trends are currently not provided by the application and the use of simple linear regression is meant for initial exploratory analysis.  The second application provides graphical summaries of water quality, weather, or nutrient station data at individual stations using the \code{plot\_summary} function: \href{https://beckmw.shinyapps.io/swmp_summary/}{https://beckmw.shinyapps.io/swmp\_summary}.  The output is identical to Figure~\ref{fig:summary_ex} with the addition of drop down menus to select the station, date range, and parameter for plotting.

\begin{figure}[!h]
\begin{center}
\includegraphics[width = \textwidth]{swmp_comp.pdf}
\caption{Online application for comparing trends in SWMP data parameters using an interactive map.  Link: \href{https://beckmw.shinyapps.io/swmp_comp}{https://beckmw.shinyapps.io/swmp\_comp}}
\label{fig:swmp_comp}
\end{center}
\end{figure}
